Quantum-assisted unsupervised data clustering

### 2020-04-29 manuscript.tex
Alexey's response

\begin{abstract}
	Unsupervised machine learning is one of the main techniques in artificial intelligence. 
	Quantum computers offer opportunities to speed up such machine learning techniques, 
	where the number of input data elements is extremely large. 
	Here, we propose a realization of quantum assisted unsupervised data clustering using the self-organizing feature map artificial neural networks. 
	We make a proof-of-concept realization of one of the central components of the approach on the IBM Q Experience 
	and show that it allows us to reduce number of calculations in a number of clusters. 
	We compare the results with the classical algorithm on a toy example of unsupervised text clustering.  
\end{abstract}

@@

Meanwhile, there has been much interest recently in applying quantum computing techniques to machine learning \cite{dunjko2018, biamonte2017, schuld2014, carleo2019}. 
The main focus of early works in quantum machine learning (QML) was on the use of quantum computers to perform basic linear algebra subroutines with a quantum speedup (such as Fourier transform and solving systems of linear equations \cite{wiebe2012,harrow2009,childs2017}) by utilizing quantum phenomena such as quantum entanglement and quantum superposition \cite{biamonte2017, schuld2014}.  
Quantum versions were developed for linear regression, principal component analysis, support vector machine, K-means algorithm and others \cite{lloyd2013,lloyd2014,dunjko2016,paparo2014,rebentrost2014}. 
More recently, there has been attention on developing quantum neural networks \cite{kamruzzaman2019, schuld2014b, jeswal2019}. 
The interest in quantum neural networks was inspired by progress in experimental quantum computing when it became possible to use parametrized quantum circuits, 
where the parameters behave much like the weights of a neural network \cite{lewenstein1994}. 
In particular, quantum algorithms for training and evaluating feedforward neural networks was developed, 
which are one of the most usable neural network models \cite{allcock2018, tacchino2019}. 
Recent results connecting the classical Bayesian approach to deep learning allowed for the development of a new algorithm for Bayesian deep learning on quantum computers \cite{zhao2019}. 
The quantum models for convolutional neural networks, which may be suitable for the problems of learning of quantum states were also proposed \cite{cong2019, liu2019}. 
Quantum classification, tested on the MNIST dataset, via slow feature analysis based on the use of quantum Frobenius distance was proposed \cite{kerenidis2018}. 
Sublinear quantum algorithms for training linear and kernel-based classifiers were designed \cite{li2019}. 
Furthermore, continuous-variable quantum neural networks \cite{killoran2019}, 
quantum autoencoders \cite{bondarenko2019}, 
quantum approaches for reinforcement learning \cite{dunjko2017, nautrup2019, foesel2018} 
and some other \cite{rebentrost2018, purushothaman1997, verdon2019, cherny2019, byrnes2013, mishra2019, vinci2019, lu2019} were developed. 
Currently, it is believed that implementation of quantum neural networks can be the main test bed to achieve practical quantum supremacy on noisy intermediate scale quantum (NISQ) devices.

@@

\begin{figure}
	\label{fig:sofm_fitting}
  	\includegraphics[width=0.95\columnwidth]{sofm_fitting.png}
	\caption{
		Schematic illustration of the clustering problem considered in this paper.  
		Blue dots represent clusters and red dots are data points. 
		The training process moves clusters to fit the data points. 
		Note that there are fewer clusters than data points, 
		which is the essence of dimensionality reduction, 
		and is what permits the model to generalize the data.
	}
\end{figure}

\section{The quantum assisted self-organizing feature map (SOFM)}
\label{sec:qasofm}

@@

The SOFM is one of the most widely-used unsupervised learning methods used in various areas of modern science. 
It was first proposed by Kohonen as a self-organizing unsupervised learning algorithm which produces feature maps similar to those occurring in the brain \cite{solan2001}. 
The SOFM algorithm operates with a set of input objects, each represented by a $N$-dimensional vector, and maps them onto nodes of a two-dimensional grid.

@@

After the winning cluster vector is selected, the weights of the vector are adjusted according to

\begin{equation}
    \label{eq:learning}
    \vec w_{\mathrm{new}} =
    	\vec w_{\mathrm{old}} 
        + \alpha\left(\vec{x} - \vec w_{\mathrm{old}}\right)  
\end{equation}

The above expression can be interpreted according to: 
if a component of the input vector is greater than the corresponding weight, 
increase the weight by a small amount; 
if the input component is smaller than the weight, decrease the weight by a small amount. 
The larger the difference between the input component and the weight component, the larger the increment (decrement). 
Intuitively, this procedure can be geometrically interpreted as iteratively moving the cluster vectors in space one at a time in a way 
that ensures each move is following the current trends inferred from their distances to the input objects. 
A visualisation of this process is shown in Fig. \ref{fig:sofm_fitting}.

@@

%%%%%
\begin{figure}[t]
	\label{fig:qcircuit}
	\includegraphics[width=0.95\columnwidth]{qcircuit.png}
	\caption{
		Quantum circuit for the quantum parallelized Hamming distance calculation between all samples and clusters. 
		First,  Hadamard gates are applied in order to obtain a superposition of the sample and cluster registers. 
		Second, we encoded information about pairwise different qubits in a quantum state of the sample register with applying the CNOT gates. 
		Third, Hamming distance values are encoded in the amplitudes of superposition with the control phase rotation and Hadamard gates. 
		Finally, a quantum state of the sample register returned to the initial basis for information retrieval. 	
	} 
\end{figure}
%%%%%

%%%%%
\begin{figure}[t]
	\label{fig:distance_matrix}
	\includegraphics[width=0.95\columnwidth]{distance_matrix.png}
	\caption{
		Hamming distance matrix between two data sets of binary vectors. 
		Values of distance that are less than the median distance value are marked black. 
		Classical simulation of the quantum circuit shows perfect agreement to theoretical calculations and presented on the left figure. 
		Result obtained on the IBM Q Experience ``ibmq\_16\_melbourne'' backend is shown on the right figure. 
		We can see good agreement between the distance matrices calculated classically and on the IBM Q Experience.  
		}
\end{figure}
%%%%%

We now introduce an optimized algorithm for calculating the matrix of Hamming distances \cite{trugenberger2001} between a sample vector and all cluster vectors, making use of quantum parallelism.  
This allows for a simple encoding of the classical information into a quantum register.

@@

Let us assume we have $k$ input vectors and $l$ cluster states. 
The $i$th input vector and $j$th cluster vector are respectively denoted as $\left| x_i \right\rangle$, $\left| y_j \right\rangle$. 
The registers $\left| X \right\rangle$ and $\left| Y \right\rangle$ are initialized to store the input vectors and cluster vectors according to
%
\begin{align}
    \left| X \right\rangle  & = \frac{1}{\sqrt{k}} \sum\limits_{i=1}^{k} \left| x_i \right\rangle,  \\
    \left| Y \right\rangle&  = \frac{1}{\sqrt{l}} \sum\limits_{j=1}^{l} \left| y_j \right\rangle .
    \label{eq:encodnig}
\end{align}
% 
The two registers along with the auxiliary qubit comprise the initial state of the quantum computer according to
%
\begin{equation} 
\left| \psi_0 \right\rangle = 
    \left| X \right\rangle
    \left| Y \right\rangle 
    \left| a \right\rangle
    \label{eq:initial_state}
\end{equation}
%
where $\left| a \right\rangle$ is an auxiliary qubit in the state $\left| 0 \right\rangle$ initially.

@@

\begin{align}
    | \psi_1 \rangle & = 
    \mathrm{CNOT} (Y,X)| \psi_0 \rangle \nonumber \\
		& =  
    \frac{1}{\sqrt{kl}} \sum_{i, j=1}^{k} 
    | d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \rangle 
    | y^{(1)}_j, \dots, y^{(n)}_j \rangle
    | 0 \rangle 
\end{align}
%
where $d^\alpha_{ij} = \mathrm{CNOT}(y^\alpha_i, x^\alpha_j)$, $\alpha = 1 \dots n$, and $i,j$ are the qubit indexes in the registers. 
At this stage of the computation the $\left| X \right\rangle$ no longer stores the input vectors,
instead it stores the information about pairwise different qubits between the input vector $\{X\}$ and cluster vector $\{Y\}$. 
Next, for each pair $\{X\}$ and $\{Y\}$, the accumulated information of all the differences is projected onto the amplitude of the superposed state. 
This is achieved by applying the Hadamard gate on auxiliary qubit, 
followed by a controlled phase gate on $\left| Xa \right\rangle$ defined as
%
\begin{equation}
    \label{eq:control_phase_rotation}
    R(\phi) = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & 1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & e^{-i\phi}
    \end{pmatrix} .
    \quad \phi = \frac{\pi}{n}
\end{equation}
%
Finally, another Hadamard gate is applied.

After the first Hadamard on the ancilla qubit the state is
%
\begin{equation}
    \left| \psi_2 \right\rangle = H_a\left| \psi_1 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
    \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle
    \dfrac{(\left| 0 \right\rangle + \left| 1 \right\rangle)}{\sqrt{2}}  .
\end{equation}
%
Applying the controlled phase gate the state then becomes: 
%
\begin{multline}
    \left| \psi_3 \right\rangle = R_{(X,a)}\left(\dfrac{\pi}{n}\right)\left| \psi_2 \right\rangle
    \\ = \dfrac{1}{\sqrt{2kl}}
				\sum\limits_{i, j=1}^{k} 
				\left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle
        \\ + \dfrac{1}{\sqrt{2kl}}
				\sum\limits_{i, j=1}^{k}
        \exp\left(\dfrac{-i \pi}{n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
\\ \times        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle
\end{multline}
%
Applying another Hadamard on the ancilla qubit we obtain
%
\begin{multline}
    \left| \psi_4 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \exp \left(\dfrac{-i \pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
		\\ \times
        \left[ \cos\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+ 
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle\right] .
\end{multline}
%
This completes the step for projecting differences between pairs of $\{X\}$ and $\{Y\}$ onto the amplitude of the auxiliary qubit. 
The process is done in the $x$-basis, achieved by the surrounding Hadamard gates. 
There are two possible measurement outcomes of the auxiliary qubit.  
Each pair of $\{X\}$ and $\{Y\}$ forms a subspace of the Hilbert space, 
the controlled phase gate ensures to change amplitudes of those outcomes within this subspace depending on how different the spin configurations between $\{X\}$ and $\{Y\}$ are.

At this stage, the information regarding the differences between pairs of $\{X\}$ and $\{Y\}$ is no longer relevant, 
thus we return to our initial basis for register $\left| X \right\rangle$ by applying pairwise CNOT gates:
%
\begin{multline}
    \left| \psi_f \right\rangle = 
    \mathrm{CNOT} (Y,X)\left| \psi_4 \right\rangle \\=  
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \exp \left(\dfrac{-i \pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
				\left[ \cos\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| X_i \right\rangle 
        \left| Y_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| X_i \right\rangle 
        \left| Y_j \right\rangle 
        \left| 1 \right\rangle\right] .
\end{multline}
%
This makes $\{X\}$ store the input vectors again, as in the initial step. 
It however preserves the amplitudes of the auxiliary qubit which are proportional to how different each pairs of $\{X\}$ and $\{Y\}$ are.

We thus have the Hamming distances encoded into the amplitudes of the final state. 
From the statistics of the measurement outcomes, the distance matrix between two data sets of binary vectors can be obtained (Fig.~\ref{fig:distance_matrix}). 
In this case, the biggest amplitude of the measurement result coincides with the smallest Hamming distance when the measurement result of the ancilla qubit is 0. 
If the ancilla qubit is 1, there is an inverse relationship with the measurement result.

Measuring the Hamming distance of a particular pair of input vectors $\left| X_i \right\rangle$ and cluster vector $\left| Y_j \right\rangle$ consists of extracting the relevant amplitude from the subspace that those states form, 
this can be done using the following projection operator
%
\begin{align}
\Pi_{i,j} = &\left| X_i \rangle\langle X_i \right| \otimes \left| X_i \rangle\langle Y_j \right| \otimes I .
\end{align} 
%
Using the above projection operator, the subspace of the Hilbert space formed by a particular pair of input and cluster vectors can be traced out as
%
\begin{align}
    \rho_{i,j} &= \text{Tr}_{1,\dots,2n} (\Pi_{i,j} \left| \psi_f \rangle\langle \psi_f \right| \Pi_{i,j}) .
\end{align}
%
From the state the following two amplitudes for the measurement results can be extracted
%
\begin{align}
    a_0(x_i,y_j) & = \frac{\left\langle 0 |\rho_{i,j}| 0 \right\rangle}{\text{Tr}(\rho_{i,j})}  \\
    a_1(x_i,y_j) & = \frac{\left\langle 1 |\rho_{i,j}| 1 \right\rangle}{\text{Tr}(\rho_{i,j})} .
\end{align}
%
In order to reduce noise we average the measurement results over different states of the ancilla qubit, 
thus the measured Hamming distance between the input vector $\left| X_i \right\rangle$ and cluster vector $\left| Y_j \right\rangle$ is
%
\begin{align}
    d_{i,j}^H & \propto 1 - \frac{1}{2}(a_0(x_i,y_j) + (1-a_1(x_i,y_j))) .
\end{align}
%
The Hamming distance measured in this way is bounded $0 \leq d_{i,j}^H \leq 1$, 
where  $0$ indicates that $x_i$ are $y_j$ identical and $1$ means they are completely opposite in terms of their pairwise binary coordinates.

In comparison with the circuits for Hamming distance calculations proposed previously \cite{trugenberger2001}, 
our method allows one to reduce the number of gates in the circuit and reconstruct all distance matrix in parallel and can be implemented with high fidelity.  
We note that at least for the $2n$ one-qubit NOT gates, 
this depends on the actual realization of the control phase gate in a quantum register at the hardware level.

%%%%%
\begin{figure}[t]
	\label{fig:vectorized_sample}
	\includegraphics[width=0.95\columnwidth]{vectorized_sample.png}
	\caption{
		(a) Representation of the data set of abstracts with the bag-of-words model is shown. 
		Each abstract is represented by a binary vector with 9 elements, corresponding to the 9 words on the horizontal axis. 
		The samples are sorted into groups (QML, MED, BIO) with 3 papers for each tag, for a total of 9 paper.    
		(b) The Hamming distance between each vectorized abstract is shown as a number in the matrix. 
	} 
\end{figure}
%%%%%

\section{Expeirmental demonstration of QASOFM}

>>>>

\begin{abstract}
	Unsupervised machine learning is one of the main techniques in artificial intelligence. 
	Quantum computers offer opportunities to speed up such machine learning techniques. 
	Here, we propose a realization of quantum assisted unsupervised data clustering using the self-organizing feature map artificial neural networks. 
	We make a proof-of-concept realization of one of the central components of the approach on the IBM Q Experience 
	and show that it allows us to reduce number of calculations in a number of clusters. 
	We compare the results with the classical algorithm on a toy example of unsupervised text clustering.  
\end{abstract}

@@

Meanwhile, there has been much interest recently in applying quantum computing techniques to machine learning \cite{dunjko2018, biamonte2017, schuld2014, carleo2019}. 
The main focus of early works in quantum machine learning (QML) was on the use of quantum computers to perform basic linear algebra subroutines with a quantum speedup (such as Fourier transform and solving systems of linear equations \cite{wiebe2012,harrow2009,childs2017}) by utilizing quantum phenomena such as quantum entanglement and quantum superposition \cite{biamonte2017, schuld2014}.  
Quantum versions were developed for linear regression, principal component analysis, support vector machine, K-means algorithm and others \cite{lloyd2013,lloyd2014,dunjko2016,paparo2014,rebentrost2014}. 
More recently, there has been attention on developing quantum neural networks \cite{kamruzzaman2019, schuld2014b, jeswal2019, broughton2020}. 
The interest in quantum neural networks was inspired by progress in experimental quantum computing when it became possible to use parametrized quantum circuits, 
where the parameters behave much like the weights of a neural network \cite{lewenstein1994}. 
In particular, quantum algorithms for training and evaluating feedforward neural networks was developed, 
which are one of the most usable neural network models \cite{allcock2018, tacchino2019}. 
Recent results connecting the classical Bayesian approach to deep learning allowed for the development of a new algorithm for Bayesian deep learning on quantum computers \cite{zhao2019}. 
The quantum models for convolutional neural networks, which may be suitable for the problems of learning of quantum states were also proposed \cite{cong2019, liu2019}. 
Quantum classification, tested on the MNIST dataset, via slow feature analysis based on the use of quantum Frobenius distance was proposed \cite{kerenidis2018}. 
Sublinear quantum algorithms for training linear and kernel-based classifiers were designed \cite{li2019}. 
Furthermore, continuous-variable quantum neural networks \cite{killoran2019}, 
quantum autoencoders \cite{bondarenko2019}, 
quantum approaches for reinforcement learning \cite{dunjko2017, nautrup2019, foesel2018} 
and some other \cite{rebentrost2018, purushothaman1997, verdon2019, cherny2019, byrnes2013, mishra2019, pyrkov2019, vinci2019, lu2019} were developed. 
At the same time, non-neural-network based hybrid quantum classical algorithms have become a new direction of significant interest \cite{mcclean2016,arute2020,akshay2020}. 
Such hybrid algorithms involve as a part some quantum circuits for speed-up and usually trained in a standard classical learning loop. 
In particular, the Quantum Approximate Optimization Algorithm was developed to find approximate solutions to combinatorial optimization problems \cite{farhi2014,farhi2016} 
and designed to problems such as MAX-CUT, Grover’s search algorithm and so on \cite{arute2020,akshay2020,wang2018,jiang2017,huang2019,wecker2016,pagano2019}. 
Another example of well-known hybrid quantum classical algorithms is Variational Quantum Eigensolver (VQE) for applications in quantum chemistry 
\cite{kandala2017,aspuru-guzik2005,lanyon2010,peruzzo2014}.
Currently, it is believed that implementation of quantum neural networks and hybrid quantum classical algorithms can be the main test bed to achieve practical quantum supremacy on noisy intermediate scale quantum (NISQ) devices.

@@

\begin{figure}
  	\includegraphics[width=0.95\columnwidth]{sofm_fitting.png}
	\caption{
		Schematic illustration of the clustering problem considered in this paper.  
		Blue dots represent clusters and red dots are data points. 
		The training process moves clusters to fit the data points. 
		Note that there are fewer clusters than data points, 
		which is the essence of dimensionality reduction, 
		and is what permits the model to generalize the data.
	}
	\label{fig:sofm_fitting}
\end{figure}

\section{The quantum assisted self-organizing feature map (QASOFM)}
\label{sec:qasofm}

@@

The SOFM is one of the most widely-used unsupervised learning methods used in various areas of modern science. 
It was first proposed by Kohonen as a self-organizing unsupervised learning algorithm which produces feature maps similar to those occurring in the brain \cite{solan2001}. 
The SOFM algorithm operates with a set of input objects, each represented by a $N$-dimensional vector, 
and describes a mapping from a higher-dimensional input space to a lower-dimensional map space.

@@

After the winning cluster vector is selected, the weights of the vector are adjusted according to
%
\begin{equation}
	\vec w_{i+1} =
		\vec w_i
		+ \alpha\left(\vec{x} - \vec w_i\right).
    \label{eq:learning}
\end{equation}
%
The above expression can be interpreted according to: 
if a component of the input vector $\vec{x}$ is greater than the corresponding weight $ \vec{w}_i $, 
increase the weight by a small amount with the learning rate $\alpha$; 
if the input component is smaller than the weight, decrease the weight by a small amount. 
The larger the difference between the input component and the weight component, the larger the increment (decrement). 
Intuitively, this procedure can be geometrically interpreted as iteratively moving the cluster vectors in space one at a time in a way 
that ensures each move is following the current trends inferred from their distances to the input objects. 
A visualisation of this process is shown in Fig. \ref{fig:sofm_fitting}.

@@

%%%%%
\begin{figure*}[t]
	\includegraphics[width=1.9\columnwidth]{qcircuit.png}
	\caption{
		Quantum circuit for the quantum parallelized Hamming distance calculation between all samples and clusters. 
		First,  Hadamard gates are applied in order to obtain a superposition of the sample and cluster registers. 
		Second, we encoded information about pairwise different qubits in a quantum state of the sample register with applying the CNOT gates. 
		Third, Hamming distance values are encoded in the amplitudes of superposition with the control phase rotation and Hadamard gates. 
		Finally, a quantum state of the sample register returned to the initial basis for information retrieval. 	
	} 
	\label{fig:qcircuit}
\end{figure*}
%%%%%

%%%%%
\begin{figure}[t]
	\includegraphics[width=0.95\columnwidth]{distance_matrix.png}
	\caption{
		(a) Hamming distance matrix between two data sets of binary vectors. 
		Values of distance that are less than the median distance value are marked black. 
		Classical simulation of the quantum circuit shows perfect agreement to theoretical calculations. 
		(b) Result obtained on the IBM Q Experience ``ibmq\_16\_melbourne'' backend. 
		We can see good agreement between the distance matrices calculated classically and on the IBM Q Experience.  
	}
	\label{fig:distance_matrix}
\end{figure}
%%%%%

We now introduce an optimized algorithm for calculating the matrix of Hamming distances \cite{trugenberger2001} between a sample vector and all cluster vectors, making use of quantum parallelism.

%This allows for a simple encoding of the classical information into a quantum register.

@@

Let us assume we have $k$ input vectors and $l$ cluster states. 
The $i$th input vector and $j$th cluster vector are respectively denoted as $\left| X_i \right\rangle$, $\left| Y_j \right\rangle$. 
The registers $\left| X \right\rangle$ and $\left| Y \right\rangle$ are initialized to store the input vectors and cluster vectors according to
%
\begin{align}
    \left| X \right\rangle  & = \frac{1}{\sqrt{k}} \sum\limits_{i=1}^{k} \left| X_i \right\rangle,  \\
    \left| Y \right\rangle&  = \frac{1}{\sqrt{l}} \sum\limits_{j=1}^{l} \left| Y_j \right\rangle .
    \label{eq:encodnig}
\end{align}
% 
The two registers along with the auxiliary qubit comprise the initial state of the quantum computer according to
%
\begin{equation} 
\left| \psi_0 \right\rangle = 
    \left| X \right\rangle
    \left| Y \right\rangle 
    \left| a \right\rangle
    \label{eq:initial_state}
\end{equation}
%
where $\left| a \right\rangle$ is an auxiliary qubit in the state $\left| 0 \right\rangle$ initially.

@@

\begin{align}
    | \psi_1 \rangle & = 
    \mathrm{CNOT} (Y,X)| \psi_0 \rangle \nonumber \\
		& =  
    \frac{1}{\sqrt{kl}} \sum_{i, j=1}^{k} 
    | d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \rangle 
    | y^{(1)}_j, \dots, y^{(n)}_j \rangle
    | 0 \rangle 
\end{align}
%
where $d^\alpha_{ij} = \mathrm{CNOT}(y^\alpha_i, x^\alpha_j)$, $\alpha = 1 \dots n$, and $i,j$ are the qubit indexes in the registers. 
At this stage of the computation the $\left| X \right\rangle$ no longer stores the input vectors,
instead it stores the information about pairwise different qubits between the input vector $\{X\}$ and cluster vector $\{Y\}$. 
Next, for each pair $\{X\}$ and $\{Y\}$, the accumulated information of all the differences is projected onto the amplitude of the superposed state. 
This is achieved by applying the Hadamard gate on auxiliary qubit, 
followed by a controlled phase gate on $\left| Xa \right\rangle$ defined as
%
\begin{equation}
    \label{eq:control_phase_rotation}
    R_{(X,a)}(\phi) = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & 1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & e^{-i\phi}
    \end{pmatrix} .
    \quad \phi = \frac{\pi}{n}
\end{equation}
%
Finally, another Hadamard gate is applied on the ancilla qubit (see Fig. \ref{fig:qcircuit}).

After the first Hadamard on the ancilla qubit the state is
%
\begin{equation}
    \left| \psi_2 \right\rangle = H_a\left| \psi_1 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
    \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle
    \dfrac{(\left| 0 \right\rangle + \left| 1 \right\rangle)}{\sqrt{2}}  .
\end{equation}
%
Applying the controlled phase gate the state then becomes
%
\begin{multline}
    \left| \psi_3 \right\rangle = R_{(X,a)}\left(\dfrac{\pi}{n}\right)\left| \psi_2 \right\rangle
    \\ = \dfrac{1}{\sqrt{2kl}}
				\sum\limits_{i, j=1}^{k} 
				\left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle
        \\ + \dfrac{1}{\sqrt{2kl}}
				\sum\limits_{i, j=1}^{k}
        \exp\left(\dfrac{-i \pi}{n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
\\ \times        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle
\end{multline}
%
Applying another Hadamard on the ancilla qubit we obtain
%
\begin{multline}
    \left| \psi_4 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \exp \left(\dfrac{-i \pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
		\\ \times
        \left[ \cos\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+ 
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \right\rangle 
        \left| y^{(1)}_j, \dots, y^{(n)}_j \right\rangle 
        \left| 1 \right\rangle\right] .
\end{multline}
%
This completes the step for projecting differences between pairs of $\{X\}$ and $\{Y\}$ onto the amplitude of the auxiliary qubit. 
The process is done in the $x$-basis, achieved by the surrounding Hadamard gates. 
There are two possible measurement outcomes of the auxiliary qubit.  
Each pair of $\{X\}$ and $\{Y\}$ forms a subspace of the Hilbert space, 
the controlled phase gate ensures to change amplitudes of those outcomes within this subspace depending on how different the spin configurations between $\{X\}$ and $\{Y\}$ are.

At this stage, the information regarding the differences between pairs of $\{X\}$ and $\{Y\}$ \hl{encoded in the amplitudes, in order to extract the Hamming distances between the relevant $\left| x_i \right\rangle$, $\left| y_j \right\rangle$ we return to our initial basis} for register $\left| X \right\rangle$ by applying pairwise CNOT gates:
%
\begin{multline}
    \left| \psi_f \right\rangle = 
    \mathrm{CNOT} (Y,X)\left| \psi_4 \right\rangle \\=  
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \exp \left(\dfrac{-i \pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
				\left[ \cos\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| X_i \right\rangle 
        \left| Y_j \right\rangle 
        \left| 0 \right\rangle\right.
        \\+
        \left. i \sin\left(\dfrac{\pi}{2n}\sum\limits_{l=1}^n d^{(l)}_{ij} \right)
        \left| X_i \right\rangle 
        \left| Y_j \right\rangle 
        \left| 1 \right\rangle\right] .
\end{multline}
%
This makes $\{X\}$ store the input vectors again, as in the initial step 
and preserves the amplitudes of the auxiliary qubit which are proportional to how different each pairs of $\{X\}$ and $\{Y\}$ are.

We thus have the Hamming distances encoded into the amplitudes of the final state. 
From the statistics of the measurement outcomes, the distance matrix between two data sets of binary vectors can be obtained (Fig.~\ref{fig:distance_matrix}). 
In this case, the biggest amplitude of the measurement result coincides with the smallest Hamming distance when the measurement result of the ancilla qubit is 0. 
If the ancilla qubit is 1, the smallest amplitude of the measurement result coincides with the smallest Hamming distance.

Measuring the Hamming distance of a particular pair of input vectors $\left| X_i \right\rangle$ and cluster vector $\left| Y_j \right\rangle$ consists of extracting the relevant amplitude from the subspace that those states form, 
this can be done using the following projection operator
%
\begin{align}
\Pi_{i,j} = &\left| X_i \rangle\langle X_i \right| \otimes \left| Y_j \rangle\langle Y_j \right| \otimes I .
\end{align} 
%
Using the above projection operator, the subspace of the Hilbert space formed by a particular pair of input and cluster vectors can be traced out as
%
\begin{align}
    \rho_{i,j} &= \text{Tr}_{1,\dots,2n} (\Pi_{i,j} \left| \psi_f \rangle\langle \psi_f \right| \Pi_{i,j}) .
\end{align}
%
From the state the following two amplitudes for the measurement results can be extracted
%
\begin{align}
    a_0(x_i,y_j) & = \frac{\left\langle 0 |\rho_{i,j}| 0 \right\rangle}{\text{Tr}(\rho_{i,j})}  \\
    a_1(x_i,y_j) & = \frac{\left\langle 1 |\rho_{i,j}| 1 \right\rangle}{\text{Tr}(\rho_{i,j})} .
\end{align}
%
In order to reduce noise we average the measurement results over different states of the ancilla qubit, 
thus the measured Hamming distance between the input vector $\left| X_i \right\rangle$ and cluster vector $\left| Y_j \right\rangle$ is
%
\begin{align}
    d_{i,j}^H & \propto 1 - \frac{1}{2}(a_0(x_i,y_j) + (1-a_1(x_i,y_j))) .
\end{align}
%
The Hamming distance measured in this way is bounded $0 \leq d_{i,j}^H \leq 1$, 
where  $0$ indicates that $x_i$ are $y_j$ identical and $1$ means they are completely opposite in terms of their pairwise binary coordinates.

In comparison with the circuits for Hamming distance calculations proposed previously \cite{trugenberger2001}, 
our method allows one to reduce the number of gates in the circuit (at least in $2n$ one-qubit NOT gates)
and reconstruct all distance matrix in parallel and can be implemented with high fidelity.

%%%%%
\begin{figure}[t]
	\includegraphics[width=0.95\columnwidth]{vectorized_sample.png}
	\caption{
		(a) Representation of the data set of abstracts with the bag-of-words model is shown. 
		Each abstract is represented by a binary vector with 9 elements, corresponding to the 9 words on the horizontal axis. 
		The samples are sorted into groups (QML, MED, BIO) with 3 papers for each tag, for a total of 9 paper.    
		(b) The Hamming distance between each vectorized abstract is shown as a number in the matrix. 
	} 
	\label{fig:vectorized_sample}
\end{figure}
%%%%%

\section{Experimental demonstration of QASOFM}

We now show experimental results for a proof-of-concept demonstration of the algorithm introduced in the previous section, on a 16-qubit quantum computer provided by IBM Q Experience.  
We perform unsupervised data clustering for three sets of paper abstracts from three different fields: 
quantum physics \cite{qml0, qml1, qml2}, 
medicine \cite{med0, med1, med2} 
and biology \cite{bio0, bio1, bio2}. 
Each set consists of three random papers that focus on one of following topics: 
``Quantum Machine Learning'' (QML), 
``Cancer'' (MED) 
and ``Gene Expression'' (BIO). 
Abstracts were vectorized by the bag-of-words model in order to choose most defining words in each data set (see Fig.~\ref{fig:vectorized_sample}) \cite{mctear2016}.  
This model represents text as a multiset ``bag'' of its words taking into account only multiplicity of words. 
Preparing the bag-of-words we excluded the words that appear only in one abstract and more than in 4 abstracts and we also excluded the word ``level'' from consideration due to the frequent overlap between the clusters because it gives instabilities for both classical and quantum algorithms. 
We restricted our bag-of-word size to 9 of the most frequent words from the full bags-of-word  due to limitations of the number of qubits. 
After vectorizing and pre-processing  the data, the clusters are well separated with the Hamming distance.  
We observe that distances between the abstracts inside clusters are smaller than distances between the abstracts from different clusters, 
showing successful self-organization (Fig.~\ref{fig:vectorized_sample}).
